---
title: "ANOVA modeling"
output: pdf_document
---

```{r, echo=FALSE}
#Read Marathon data dile
#fname = file.choose() 
fname <- "/Users/poojasingh/Documents/HStatE139/git/statse139-project2/statse139-project/Previous Boston Marathon study/BAA data.txt"
dfm <- read.csv(fname, header=T,sep=" ")
sink("/dev/null") 
#names(dfm)
#head(dfm)

#Histogram of Total Time - is it skewed?
#hist(dfm.2010$TotalTime)
#hist(log(dfm.2010$TotalTime))

#hist(dfm.2010$K0.5)
#hist(log(dfm.2010$K0.5))
sink() 

#Log transform split-time data
dfm <- cbind(dfm[,1:6], log(dfm[7]), log(dfm[8]), log(dfm[9]), log(dfm[10]), log(dfm[11]),log(dfm[12]),log(dfm[13]),log(dfm[14]),+
               log(dfm[15]), dfm[,16:18])

#Code Gender1F2M as a factor to treat it as a categorical variable
dfm$Gender <- factor(dfm$Gender1F2M, labels=c("female", "male"))
#levels(dfm$Gender)

#Create factor for agegroup
dfm$AgeGroup <- cut(dfm$Age, breaks=c(15, 35, 55, 85))
#levels(dfm$AgeGroup) #Levels "(15,35]" "(35,55]" "(55,85]"

#Calculate Total time 
times <- as.matrix(dfm[,7:15], ncol=9)
dfm$TotalTime <- rowSums(times)

#Calculate first half 
firstHalf <- cbind(dfm[,7:10],(dfm[11]/5)*1.1)
dfm$FirstHalf <- rowSums(firstHalf)

#Calculate 2nd half
secondHalf <- cbind((dfm[11]/5)*3.9, dfm[12:15])
dfm$SecondHalf <- rowSums(secondHalf)

#Split data by year
dfm.2010 <- subset(dfm, (dfm$Year==2010))
sink("/dev/null") 
attach(dfm.2010)
sink() 
#unique(Year) #Test - returns Year = 2010 only

#Split data by gender
dfm.2010.male <- subset(dfm, (Gender1F2M==2))
dfm.2010.female <- subset(dfm, (Gender1F2M==1))

```
Let's look at the ANOVA model to examine whether average time for men and women run are different for first K0-5 split time for 2010. (All the times are presented on the log scale to account for outliers and slight right skewness in the data.)
```{r}
#1. Do men and women run differently first K0-5 - let's look at the ANOVA model
boxplot(K0.5 ~ Gender, main="K0-5 Split time for '10")

#ANOVA By Gender - first K0-5
model1 <- aov(K0.5 ~ Gender)
```
The summary of anova model is presented below which indicates significant result as p-value < 0.05 and  we reject the null hypothesis and conclude that men and women have different average time for first K0-5 split time.
```{r}
summary(model1)
```
\newpage
Similarly, we examine the total runnnig time for men and women.
```{r}
#2. Do men and women total time are different - let's look at the ANOVA model
boxplot(TotalTime ~ Gender, main="Total Time for running marathon")
```
\newline
From the p-value for the ANOVA test we reject the null hypothesis and accept the alternate hypothesis that men and women have different average total time.
```{r}
#ANOVA TotalTime By Gender
model2 <- aov(TotalTime ~ Gender)
summary(model2)
```
Next, we examine whether men and women run differently by age group by comparing the total time in three different age groups, namely, (15,35], (35,55] and (55,85]. From the anova summay we conclude that average time is different among the age groups for first K0-5 split time.
\newpage
```{r}
#3. Do men and women run differently by age group - let's look at the ANOVA model
boxplot(K0.5 ~ AgeGroup, main="K0-5 vs. Age group")

#ANOVA By Age group
model3 <- aov(K0.5 ~ AgeGroup)
summary(model3)
```
Now, lets examine the interaction effect of age and gender together. We also run the TukeyHSD model and find that for first K0-5 split time, female in age group (35,55] and male in age group (55,85] do not have signigficant result. However this is not the same when we run the Tukey test for total time as seen below. Highlighted the main TukeyHSD output below,
\newline
## TukeyHSD(aov(K0.5 ~ AgeGroup * Gender)) 
\newline
## Fit: aov(formula = K0.5 ~ AgeGroup * Gender)
\newline
## $`AgeGroup:Gender`
\newline
##                               p adj
\newline
## (55,85]:male-(35,55]:female   0.761009
\newline
```{r, echo=FALSE, message=FALSE, render=FALSE}
#4. Do men and women run different by age group and gender
#ANOVA By Age group and Gender - interaction effect
model5 <- aov(K0.5 ~ AgeGroup * Gender)
summary(model5)

#5. TukeyHSD test for K0-5 and Age group - all pairwise tests are significant
#TukeyHSD(aov(K0.5 ~ AgeGroup * Gender)) #Look out for one pair. Same results are not obtained for TotalTime

#TukeyHSD(aov(TotalTime ~ AgeGroup * Gender)) #All are significant
```

\newpage
We also compared fastest 2000 runners for male and female group and found that the average time for first K0-5 split time is different for men and women.
\newline
```{r, echo=FALSE, message=FALSE}
#6a. Top 2000 male runners in 2010
dfm.2010.male.top2000 <- dfm.2010.male[order(dfm.2010.male$K0.5, decreasing=FALSE)[1:2000],]

#6b. Top 10% female 
dfm.2010.female.top2000 <- dfm.2010.female[order(dfm.2010.female$K0.5, decreasing=FALSE)[1:2000],]

#6c. Do top 2000 men and women run same by gender
top2000 <- rbind(dfm.2010.female.top2000, dfm.2010.male.top2000)
boxplot(top2000$K0.5 ~ top2000$Gender, main="Total time vs. Gender for top 2000 fastest runners",  ylab="Total time in minutes")
```

```{r}
model6 <- aov(K0.5 ~ Gender)
summary(model6)
```

```{r, echo=FALSE, message=FALSE}
#7a. First Half-time vs 2nd half-time
model7 <- aov(TotalTime ~ FirstHalf)
summ <- summary(model7)
#summ
#summary(lm(TotalTime ~ FirstHalf)) #check R^2 values here

r2.model7 <- 1 - (summ[[1]]$'Sum Sq'[[2]]/sum(summ[[1]]$'Sum Sq'))
#r2.model7

model7b <- aov(TotalTime ~ SecondHalf)
summ <- summary(model7b)
#summary(lm(TotalTime ~ SecondHalf)) #check R^2 values here

r2.model7b <- 1 - (summ[[1]]$'Sum Sq'[[2]]/sum(summ[[1]]$'Sum Sq'))
#r2.model7b

model7c <- aov(TotalTime ~ FirstHalf + Age + Gender)
summ <- summary(model7c)
#summary(lm(TotalTime ~ FirstHalf + Age + Gender)) #check R^2 values here

r2.model7c <- 1 - (summ[[1]]$'Sum Sq'[[4]]/sum(summ[[1]]$'Sum Sq'))
#r2.model7c

model7d <- aov(TotalTime ~ K0.5)
summ <- summary(model7d)
#summary(lm(TotalTime ~ K0.5)) #check R^2 values here

r2.model7d <- 1 - (summ[[1]]$'Sum Sq'[[2]]/sum(summ[[1]]$'Sum Sq'))
#r2.model7d

results <- matrix(c("aov(TotalTime~SecondHalf)", r2.model7b, "aov(TotalTime~FirstHalf+Age+Gender)", r2.model7c, "aov(TotalTime~FirstHalf)",r2.model7, "aov(TotalTime~K0.5)", r2.model7d),4,2,byrow=TRUE)
colnames(results) <- c("model", "R^2")
#r2.model7b #aov(TotalTime ~ SecondHalf) -> 0.9594059
#r2.model7c #aov(TotalTime ~ FirstHalf + Age + Gender) -> 0.9233338
#r2.model7 #aov(TotalTime ~ FirstHalf) -> 0.9203939
#r2.model7d #aov(TotalTime ~ K0.5) -> 0.8419144
```
The various ANOVA models for TotalTime were analyzed and summarized below with the decreasing R^2 values. 
```{r}
results
```

We still have to analyze ANOVA in for interaction effect of both Age and Gender on the Split time...

