---
title: "anova_splittimes_Tukey.R"
output: pdf_document
---
Perform analysis of variance test to verify if split times vary between age groups. 

To begin, we create the factors for agegroup, namely divided the Age into following intervals: \newline
"(15,25]" "(25,35]" "(35,45]" "(45,55]" "(55,65]" "(65,75]" "(75,85]"


```{r, , echo=FALSE}
fname = "/Users/poojasingh/Documents/HStatE139/git/statse139-project2/statse139-project/Previous Boston Marathon study/BAA data.txt"
dfm = read.csv(fname, header=T,sep=" ")
#names(dfm)

#Create factor for gender
#factor(dfm$Gender1F2M, ordered=TRUE) #identify sorted levels in input source
dfm$gender = factor(dfm$Gender1F2M, labels=c("female", "male")) 
#levels(dfm$gender) #output levels

#Create factor for agegroup
#factor(dfm$Age, ordered=TRUE)  #identify sorted levels in input source
dfm$agegroup = cut(dfm$Age, breaks=c(15, 25, 35, 45, 55, 65, 75, 85))
#levels(dfm$agegroup) #output levels

#Split data by year
dfm.2010 = subset(dfm, (dfm$Year==2010))
```

Let's perform ANOVA for K0-5 splittime and different age groups. 
```{r}
#Anova by agegroup
model1 = aov(dfm.2010$K0.5 ~ dfm.2010$agegroup)
summary(model1)
```

Perform a TukeyHSD for multiple comparision test to find whether variances for splittime are different amoung age groups.  

```{r}
#Conduct Tukey's multiple comparision test
TukeyHSD(model1)
```

From TukeyHSD results, we can conclude that differences are not significant among 
age group (15,25] - (25,35]  as p-value = 0.6934223.

Similarly, age groups (15,25] - (35,45] and (65,75] - (75,85]  do not have significant results as p-values are greater than 0.05.

