---
title: "predictFinish"
author: "David Wihl"
date: "November 24, 2015"
output: pdf_document
---

First let's read in the data and calculate finish times:

```{r}
dfm <- read.csv("Previous Boston Marathon study/BAA data.txt",header=T,sep=" ")
times = as.matrix(dfm[,7:15], ncol=9)
dfm$totaltime = rowSums(times)
```

Some preliminary EDA of the data:
```{r}
hist(dfm$totaltime,breaks=50, main="Boston Marathon Finish Time (min) Distribution for '10, '11, '13")
```

The data appears somewhat left skewed. Let's try a log transform:

```{r}
hist(log(dfm$totaltime),breaks=50, main="Boston Marathon Finish Time (log-min) Distribution for '10, '11, '13")
```

That seems better.

Let's try a multiple regression on the non-log data:

```{r}
dfms = dfm[!is.na(dfm$totaltime), ]
dfms = dfms[order(dfms$totaltime),]
model = lm(totaltime~Age+Gender1F2M+K0.5, data=dfms)
summary(model)
model.resid = resid(model)
```

Let's plot the residuals
```{r}
plot(dfms$totaltime,model.resid,ylim=c(-100,100))
```

This appears to be spreading out and trending up. 

Let's try a polynomial transformation on the 5k time to see if that improves things:
```{r}
bestresidsum = 9e9
bestpoly = 0
for (i in seq(from=1.0, to= 2.5, by=0.1)){
  dfms$k5xform = dfms$K0.5 ^ i
  model = lm(totaltime~Age+Gender1F2M+k5xform, data=dfms)
  model.residsum = sum(resid(model)^2)
  if (model.residsum < bestresidsum) {
    bestresid = resid(model)
    bestresidsum = model.residsum
    bestpoly = i
  }
}

cat("total error went from ",sum(model.resid^2), " (untransformed) to ", bestresidsum," (transformed)\n")
plot(dfms$totaltime,bestresid,ylim=c(-100,100))
```
